Pillar 1 - Sequential Pattern Mining

### 2019-01-24 Pillar 1 - Sequential Pattern Mining.Rmd
cleans code- breaks into smaller section and puts code in correct order

# Setup
load previous data and create function to find courses
```{r}
load("data_pillar_1.RDATA")

@@

#Data Exploration of Transcripts
```{r}
exploration <- d_transcript %>%
  mutate(Grade = case_when(Grade == "NG"~ "-1", #"0,0", #maybe we could just remove these from the analysis and treat them separately?
                           T~Grade),
         Grade = str_replace(string = Grade, pattern = ",",replace = "."),
         Grade = as.numeric(Grade))%>%
  filter(!is.na(Grade)) %>%
  filter(Grade != -1) #remove NG

@@

#Getting dates
For a first exploration of arules, we conceptualise our framework like this:
transaction = student
item = course

##Data preparation
First we transform our data into transaction data. For this, we first create a vector of mandatory courses that we exclude from transcripts.

@@

#Mining-sequences:
The parameter values are: 
support:  a numeric value specifying the minimum support of a sequence (default 0.1, range [0,1]).

maxsize:  an integer value specifying the maximum number of items of an element of a sequence (default 10, range > 0).
maxlen:   an integer value specifying the maximum number of elements of a sequence (default 10, range > 0).
mingap:   an integer value specifying the minimum time difference between consecutive elements of a sequence (default none, range >= 1).
maxgap:   an integer value specifying the maximum time difference between consecutive elements of a sequence (default none, range >= 0).
maxwin:   an integer value specifying the maximum time difference between any two elements of a sequence (default none, range >= 0).

#Preparing data frame:
```{r}
fail_threshold <- 6.5

@@

  left_join(d_transcript, by = "Course ID") %>%

    # mutate(
    #   Grade = case_when(
    #     Grade == "NG"~ "-1",
    #     T            ~ Grade),
    #   Grade = str_replace(string = Grade, pattern = ",",replace = "."),
    #   Grade = as.numeric(Grade)) %>% #this was already done in data preparation

  # filter(!is.na(Grade)) %>% #remove NaN
  # filter(Grade != -1) %>% #remove NG
  mutate(
    Outcome = 
           case_when( Grade == -1   ~"Fail", #if we don't do this NG satisfies the condition for pass v
                      Grade  > 5.5 ~"Pass",
                               T~"Fail"),
    Low_grade =
           case_when(Grade == -1 ~"low score", #if we don't do this NG satisfies the condition for pass v
                      Grade > fail_threshold ~"high score",
                               T~"low score"),
    Discrete_grade =
           case_when(Grade <= 5.4             ~ 5.0,
                     between(Grade, 5.5, 6.4) ~ 6.0,
                     between(Grade, 6.5, 7.4) ~ 7.0,
                     between(Grade, 7.5, 8.4) ~ 8.0,
                     between(Grade, 8.5, 9.4) ~ 9.0,
                     Grade >= 9.5             ~ 10.0,
                     T ~ -99999    #this is a control, we should not have any grades here
                     ),
    None= "" # this might not be optimal
         ) %>%

@@

##Rule induction
```{r}
#FURTURE: create filter by function
string_filter <- case_when(quo_name(category)=="Outcome"        ~ "Fail",
                           quo_name(category)=="Low_grade"      ~ "low score",
                           quo_name(category)=="Discrete_grade" ~ "5",
                           quo_name(category)=="None"           ~ "_" #useless filter
                           )

frequent_trajectories <- cspade(transactions, parameter = list(support = 0,
                                                               maxsize = 1,
                                                               maxlen = 2), control   = list(verbose = TRUE))

summary(frequent_trajectories)
as(frequent_trajectories, "data.frame")

rules <- ruleInduction(frequent_trajectories,
                       confidence = 0, #we need to play with the parameters.
                       control    = list(verbose = TRUE))

n_students <- length(unique(d_sequence_transactions$`Student ID`))

  # Filter and edit output
  filter(str_detect(`RHS Score`, pattern = "low"),
         str_detect(`LHS Score`, pattern = "low"),
         n_rule >= 5,
         Conf_corr > 0.2) %>%
  select(LHS, RHS, support, n_rule, Prob_both, n_both, Prop_low, Conf_corr, lift_corr, lift_diff) %>%
  mutate(support = round(support,3),
         Prob_both =round(Prob_both,3),
         Prop_low =round(Prop_low,3),
         Conf_corr= round(Conf_corr,3),
         lift_diff= round(lift_diff,3),
         lift_corr= round(lift_corr,3)) %>%
  arrange(lift_diff)

#FOR PLOTTING
for_plots <-  manual_confidence_rules %>%
  group_by(LHS)%>%
  mutate(LHS_count=n())

```
#Visualizing rules
```{r}
#look at counts
ggplot(for_plots, aes(LHS))+
  geom_bar()

hi <- table(select(manual_confidence_rules,LHS,RHS))
#ggnetwork

#Visualisation
It might be a nice idea to visualise using Sankey diagrams, although I don't know if there will be too many courses for it to be meaningful. We might need to sample. See https://analyzecore.com/2014/10/31/sequence-carts-analysis-sankey/

#Experimenting with transactions
```{r}
fail_threshold <- 6.5

lookup_table <- d_course %>%

@@

  summarize(Count = n()) %>%
  #separate(Item, into= c("Course ID", quo_name(category)), sep="_", remove = F) %>%
  ungroup() %>%
  group_by(`Course ID`) %>%
  mutate(Total= sum(Count))%>%
  filter(Low_grade=="low score")%>%
  mutate(Prop_low= Count/Total)

lookup_table
```

@@

```{r}

length(unique(d_transcript$`Student ID`))
```

>>>>

# Setup
load previous data and create function to find courses
```{r loading data}
load("data_pillar_1.RDATA")

@@

```
#Simple Rule Mining
##Data Exploration of Transcripts
```{r}
exploration <- d_transcript %>%
  mutate(Grade = case_when(Grade == "NG"~ "-1", #"0,0", #maybe we could just remove these from the analysis and treat them separately?
                           T~Grade),
         Grade = str_replace(string = Grade, pattern = ",",replace = "."),
         Grade = as.numeric(Grade))%>%
  filter(!is.na(Grade)) %>%
  filter(Grade != -1) #remove NG

@@

##Getting dates
For a first exploration of arules, we conceptualise our framework like this:
transaction = student
item = course

###Data preparation
First we transform our data into transaction data. For this, we first create a vector of mandatory courses that we exclude from transcripts.

@@

#Mining-sequences:
The parameter values are: 
support:  a numeric value specifying the minimum support of a sequence (default 0.1, range [0,1]).
maxsize:  an integer value specifying the maximum number of items of an element of a sequence (default 10, range > 0).
maxlen:   an integer value specifying the maximum number of elements of a sequence (default 10, range > 0).
mingap:   an integer value specifying the minimum time difference between consecutive elements of a sequence (default none, range >= 1).
maxgap:   an integer value specifying the maximum time difference between consecutive elements of a sequence (default none, range >= 0).
maxwin:   an integer value specifying the maximum time difference between any two elements of a sequence (default none, range >= 0).

##Preparing data frame:
```{r}
fail_threshold <- 6.5

@@

  left_join(d_transcript, by = "Course ID") %>%
  # filter(!is.na(Grade)) %>% #remove NaN
  # filter(Grade != -1) %>% #remove NG
  mutate(
    Outcome = 
           case_when( Grade == -1   ~"Fail", #if we don't do this NG satisfies the condition for pass v
                      Grade  > 5.5 ~"Pass",
                               T~"Fail"),
    Low_grade =
           case_when(Grade == -1 ~"low score", #if we don't do this NG satisfies the condition for pass v
                      Grade > fail_threshold ~"high score",
                               T~"low score"),
    Discrete_grade =
           case_when(Grade <= 5.4             ~ 5.0,
                     between(Grade, 5.5, 6.4) ~ 6.0,
                     between(Grade, 6.5, 7.4) ~ 7.0,
                     between(Grade, 7.5, 8.4) ~ 8.0,
                     between(Grade, 8.5, 9.4) ~ 9.0,
                     Grade >= 9.5             ~ 10.0,
                     T ~ -99999    #this is a control, we should not have any grades here
                     ),
    None= "" # this might not be optimal
         ) %>%

@@

##Rule induction
##Creating look-up table
```{r}
lookup_table <- d_course %>%

@@

  summarize(Count = n()) %>%
  #separate(Item, into= c("Course ID", quo_name(category)), sep="_", remove = F) %>%
  ungroup() %>%
  group_by(`Course ID`) %>%
  mutate(Total= sum(Count))%>%
  filter(Low_grade=="low score")%>%
  mutate(Prop_low= Count/Total)
```

###NOTE for FURTURE: create filter by function
##Create filter condition.
```{r}
string_filter <- case_when(quo_name(category)=="Outcome"        ~ "Fail",
                           quo_name(category)=="Low_grade"      ~ "low score",
                           quo_name(category)=="Discrete_grade" ~ "5",
                           quo_name(category)=="None"           ~ "_" #useless filter
                           )
```
##CSPADE Algorithm:
```{r}
frequent_trajectories <- cspade(transactions, parameter = list(support = 0,
                                                               maxsize = 1,
                                                               maxlen = 2), control   = list(verbose = TRUE))

#Inspect results CSPADE
summary(frequent_trajectories)
as(frequent_trajectories, "data.frame")

```
##rule induction:
```{r}
rules <- ruleInduction(frequent_trajectories,
                       confidence = 0, #we need to play with the parameters.
                       control    = list(verbose = TRUE))

```
##calculate appropriate measures:
```{r}
n_students <- length(unique(d_sequence_transactions$`Student ID`))

  # Filter and edit output
  filter(str_detect(`RHS Score`, pattern = "low"),
         str_detect(`LHS Score`, pattern = "low"),
         n_rule >= 5,
         Conf_corr > 0.2) %>%
  select(LHS, RHS, support, n_rule, Prob_both, n_both, Prop_low, Conf_corr, lift_corr, lift_diff) %>%
  mutate(support = round(support,3),
         Prob_both =round(Prob_both,3),
         Prop_low =round(Prop_low,3),
         Conf_corr= round(Conf_corr,3),
         lift_diff= round(lift_diff,3),
         lift_corr= round(lift_corr,3)) %>%
  arrange(lift_diff)
#Inspect results
manual_confidence_rules
```
##Arrange data for plots:
```{r}
for_plots <-  manual_confidence_rules %>%
  group_by(LHS)%>%
  mutate(LHS_count=n())
```
#Visualizing rules
```{r}
#look at counts
ggplot(for_plots, aes(LHS))+
  geom_bar()

#ggnetwork

#SEE: http://kateto.net/network-visualization 
g_net <- graph_from_data_frame(d=head(for_plots,50), directed=T) #I BELIEVE THE COLUMNS SHOULD BE REVERSED.

l <- layout_in_circle(g_net)
plot(g_net, #vertex.shape="none",
     vertex.size=.9,
     #vertex.label=NA,
     vertex.label.cex=0.5,
     edge.arrow.size=.1, 
     edge.curved=.1,
     layout=l)

```
#Visualisation
It might be a nice idea to visualise using Sankey diagrams, although I don't know if there will be too many courses for it to be meaningful. We might need to sample. See https://analyzecore.com/2014/10/31/sequence-carts-analysis-sankey/

@@
